---
title: "S0 — Environment Check"
subtitle: "R Nowcasting Training — IDB Caribbean Country Department"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: github
    self_contained: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# What this file does

This notebook verifies that your R environment is correctly configured for the training. It checks four things, in order:

| Step | What we check |
|---|---|
| 1. VS Code setup | R extension is active and running |
| 2. Rtools | C++ build toolchain is available (required for `mfbvar`) |
| 3. Packages | All required libraries load without errors |
| 4. Smoke tests | Each model runs on synthetic data and returns a forecast |

**How to run it:** Execute each chunk in order using the **▶ Run Cell** button or `Ctrl+Enter`. Expected output is shown below each chunk. If anything fails, read the note in that section before continuing.

---

# Before you start — VS Code setup

Before running any code, confirm the following in VS Code:

**1. The R extension is installed**
Open the Extensions panel (`Ctrl+Shift+X`) and search for **"R"** by REditorSupport. It should show as installed and enabled.

**2. A CRAN mirror is selected**
The first time you install an R package in VS Code, the terminal will ask:

```
Would you like to choose a CRAN mirror for this session? (Yes/no/cancel)
```

Type `yes` and press Enter. A numbered list of mirrors will appear. Type `0` (for **0-Cloud**, the globally load-balanced default) and press Enter. This only needs to be done once per session — you will not be asked again.

**3. The R terminal is active**
You should see an `R` prompt (`>`) in the VS Code terminal panel at the bottom. If you see a plain shell prompt instead, run `R` in the terminal to start an R session.

---

# 1 · Rtools check

`mfbvar` must be compiled from source, which requires Rtools — a Windows build toolchain that is separate from R itself. This chunk verifies it is installed and reachable.

```{r rtools-check}
if (!requireNamespace("pkgbuild", quietly = TRUE)) {
  install.packages("pkgbuild")
}

if (pkgbuild::has_build_tools(debug = FALSE)) {
  cat("Rtools OK — build tools found\n")
} else {
  cat("Rtools NOT found — mfbvar cannot be installed without it.\n")
  cat("See the mfbvar troubleshooting guide for installation instructions.\n")
}
```

**Expected output:**
```
Rtools OK — build tools found
```

> If you see `Rtools NOT found`, stop here and follow the mfbvar troubleshooting guide before continuing.

---

# 2 · Package installation

This chunk installs any packages that are not yet present. If everything is already installed, it will complete silently in a few seconds. `mfbvar` is installed from GitHub — this is the step that requires Rtools.

```{r install}
cran_packages <- c(
  "tidyverse",
  "lubridate",
  "openxlsx",
  "Rmisc",
  "nowcastDFM",
  "imputeTS",
  "midasr",
  "devtools",
  "jsonlite",
  "pkgbuild"
)

# Install any missing CRAN packages
missing <- cran_packages[!sapply(cran_packages, requireNamespace, quietly = TRUE)]
if (length(missing) > 0) {
  cat("Installing:", paste(missing, collapse = ", "), "\n")
  install.packages(missing)
} else {
  cat("All CRAN packages already installed\n")
}

# Install mfbvar from GitHub if not present
if (!requireNamespace("mfbvar", quietly = TRUE)) {
  cat("Installing mfbvar from GitHub...\n")
  devtools::install_github("ankargren/mfbvar")
} else {
  cat("mfbvar already installed\n")
}
```

---

# 3 · Library check

This chunk loads every package used across both training scripts. A clean run with no red error messages means your environment is complete.

```{r libraries}
library(tidyverse)    # data manipulation and plotting
library(lubridate)    # date arithmetic
library(openxlsx)     # Excel import/export (Session 2b)
library(Rmisc)        # CI(): confidence intervals for the BVAR prior
library(nowcastDFM)   # dfm(), predict_dfm(): Dynamic Factor Model
library(imputeTS)     # na_mean(): missing value imputation
library(midasr)       # midas_r(), mls(): MIDAS regression
library(mfbvar)       # set_prior(), estimate_mfbvar(): MF-BVAR
library(jsonlite)     # JSON parsing utilities
library(devtools)     # GitHub package installation

cat("All libraries loaded OK\n")
```

**Expected output:**
```
All libraries loaded OK
```

> **Note on `jsonlite`:** In some VS Code configurations a warning about `jsonlite` appears when starting R. This does not affect any part of the training workflow. Installing it explicitly (done above) resolves the warning in most cases. If it persists, ignore it.

---

# 4 · Smoke tests

Each model is fitted on the same small synthetic dataset defined below. The goal is not to produce meaningful forecasts — it is to confirm that the estimation machinery runs end to end on your machine.

## 4.0 Synthetic data

Three series mimicking the Barbados nowcasting setup: quarterly real GDP (`gdp_brb`) and two monthly indicators (`x1_brb`, `x2_brb`). All values are randomly generated — only the structure (frequencies, date range, ragged edge) matches the real data.

```{r synthetic-data}
set.seed(42)

# Date range: 2010 Q1 – 2024 Q4 (60 quarters, 180 months)
dates_q <- seq(as.Date("2010-01-01"), by = "quarter", length.out = 60)
dates_m <- seq(as.Date("2010-01-01"), by = "month",   length.out = 180)

# Quarterly GDP (Barbados-style: cumulative growth path)
gdp_brb <- cumsum(rnorm(60, mean = 0.4, sd = 1.2))

# Monthly indicators (correlated with GDP via shared factor)
factor  <- rep(gdp_brb, each = 3) + rnorm(180, sd = 0.5)
x1_brb  <- factor + rnorm(180, sd = 0.8)   # e.g. tourist arrivals proxy
x2_brb  <- factor + rnorm(180, sd = 1.1)   # e.g. exports proxy

# Combined monthly dataframe (quarterly GDP repeated at monthly frequency)
df_smoke <- data.frame(
  date    = dates_m,
  gdp_brb = rep(gdp_brb, each = 3),   # quarterly values held constant within quarter
  x1_brb  = x1_brb,
  x2_brb  = x2_brb
)

cat(sprintf("Synthetic data OK — %d monthly rows, %d quarterly obs, 3 series\n",
            nrow(df_smoke), length(gdp_brb)))
head(df_smoke)
```

**Expected output:**
```
Synthetic data OK — 180 monthly rows, 60 quarterly obs, 3 series
```

---

## 4.1 MIDAS

```{r smoke-midas}
# mls(x, 0:3, 3) uses 4 monthly lags at ratio 3 = 12 monthly obs consumed.
# We align by trimming: keep quarters 5:60 for y, months 1:168 for x
# (168 = 56 quarters * 3 months), so both vectors have length 56.
y_midas <- gdp_brb[5:60]        # 56 quarterly obs
x_midas <- x1_brb[1:168]        # 56 * 3 = 168 monthly obs

model_midas <- midas_r(
  y_midas ~ mls(x_midas, 0:3, 3, nealmon),
  start = list(x_midas = c(1, -0.5))
)

fcst_midas <- tail(fitted(model_midas), 1)
cat(sprintf("MIDAS OK — in-sample fit (last obs): %.4f\n", fcst_midas))
```

**Expected output:**
```
MIDAS OK — in-sample fit (last obs): [number]
```

## 4.2 MF-BVAR

```{r smoke-bvar}
# mfbvar requires ts objects: monthly series first, quarterly last
x1_ts  <- ts(x1_brb, start = c(2010, 1), frequency = 12)
x2_ts  <- ts(x2_brb, start = c(2010, 1), frequency = 12)
gdp_ts <- ts(gdp_brb, start = c(2010, 1), frequency = 4)

mf_list <- list(x1_brb = x1_ts, x2_brb = x2_ts, gdp_brb = gdp_ts)

prior <- set_prior(Y = mf_list, n_lags = 3, n_reps = 50,
                   block_exo = "gdp_brb")

c_interval      <- t(sapply(mf_list,
                   function(s) CI(na.omit(as.numeric(s)), ci = 0.95)))
prior_intervals <- c_interval[, c("upper", "lower")]
moments         <- interval_to_moments(prior_intervals)

prior <- update_prior(prior,
                      d               = "intercept",
                      prior_psi_mean  = moments$prior_psi_mean,
                      prior_psi_Omega = moments$prior_psi_Omega)
prior <- update_prior(prior, n_fcst = 4)

model_bvar <- estimate_mfbvar(prior, prior = "minn", variance = "iw")

fcst_bvar <- predict(model_bvar, pred_bands = NULL) %>%
  dplyr::filter(variable == "gdp_brb") %>%
  pull(fcst) %>%
  mean()

cat(sprintf("BVAR OK — mean forecast across draws: %.4f\n", fcst_bvar))
```

**Expected output:**
```
BVAR OK — mean forecast across draws: [number]
```

# Summary

Run this chunk last. It gives a single-glance overview of what passed.

```{r summary}
results <- list(
  "Rtools"    = pkgbuild::has_build_tools(debug = FALSE),
  "Libraries" = all(sapply(c("tidyverse","lubridate","openxlsx","Rmisc",
                              "nowcastDFM","imputeTS","midasr","mfbvar",
                              "jsonlite"), requireNamespace, quietly = TRUE)),
  "MIDAS"     = exists("fcst_midas"),
  "BVAR"      = exists("fcst_bvar")
)

cat("=== Environment Check Summary ===\n")
for (item in names(results)) {
  status <- if (isTRUE(results[[item]])) "OK" else "FAILED"
  cat(sprintf("  %-12s %s\n", item, status))
}
cat("=================================\n")
```

**Expected output:**
```
=== Environment Check Summary ===
  Rtools       OK
  Libraries    OK
  MIDAS        OK
  BVAR         OK
=================================
```

> If any line shows `FAILED`, scroll up to the corresponding section and read the note. All five must show `OK` before starting the training exercises.
